---
title: "What a start-site switch does to the transcript"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What a start-site switch does to the transcript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.6, fig.height = 5.0, dpi = 120
)
```

```{r setup}
library(omakase)
```

## The question

A gene fires from two promoters. The browser shows two transcripts that start in
different places — but what does that *do*? Sometimes nothing to the protein:
the coding sequence is identical and only the 5′ UTR changed, which is a
regulatory event. Sometimes the protein itself gains or loses an N-terminus.
Those are different biological claims, and a picture of the two transcripts does
not distinguish them.

`classify_consequence()` answers the question from the open reading frames.

## The decision tree

Write $P_m$ and $P_a$ for the peptides the reference and alternative isoforms
encode, and $\ell_m$, $\ell_a$ for their 5′ UTR lengths.

1. **One isoform codes and the other does not.** $P_m \neq \varnothing,
   P_a = \varnothing$ is an *ORF loss*; the reverse is an *ORF gain*. If
   neither codes, the pair is unclassified — blaming the start site for an ORF
   that was never there would be wrong.

2. **The protein is untouched** ($P_a = P_m$). The change is purely 5′. If the
   two first exons are disjoint *and* the alternative site is downstream, the
   gene has switched promoters outright: *promoter swap*. Otherwise it is a
   *5′ UTR change*, `longer`, `shorter` or `equal` by the sign of
   $\ell_a - \ell_m$.

3. **The proteins differ.** If $P_m$ ends with $P_a$, the alternative protein is
   the reference one with its N-terminus removed: *N-terminal truncation*. If
   $P_a$ ends with $P_m$, it is an *extension*. Otherwise the N-terminus is
   simply different: *alternative N-terminus*.

The order matters. A 5′ UTR change wins over everything that looks dramatic in
the genome browser, because if the protein is identical the consequence is
regulatory however far apart the two start sites are.

### The distal guard

Before any of that, the reciprocal overlap of the two transcript spans is
checked:

$$o = \frac{\max\bigl(0,\ \min(e_1, e_2) - \max(s_1, s_2)\bigr)}{\min(e_1 - s_1,\ e_2 - s_2)}$$

When $o = 0$ the two "isoforms" do not overlap at all. That is not one
transcription unit with two start sites, it is two separate transcripts sharing
a gene label, and such pairs are marked `"Distal (excluded)"` rather than
classified.

## Classifying

The input is one row per switch, describing both isoforms:

```{r}
switches <- data.frame(
  gene_name = c("Reg1", "Swap1", "Ext1", "Trunc1", "Dead1"),
  strand = "+",
  main_protein = c("MKVLAAGIV", "MKVLAAGIV", "MKVLAAGIV", "MKVLAAGIV", "MKVLAAGIV"),
  alt_protein  = c("MKVLAAGIV", "MKVLAAGIV", "MQQPMKVLAAGIV", "AAGIV", ""),
  main_utr5_len = c(100, 100, 100, 100, 100),
  alt_utr5_len  = c(350, 100, 100, 100, 100),
  main_first_exon_start = 1000, main_first_exon_end = 1200,
  alt_first_exon_start  = c(1050, 9000, 1050, 1050, 1050),
  alt_first_exon_end    = c(1250, 9200, 1250, 1250, 1250),
  main_tss = 1000,
  alt_tss  = c(1050, 9000, 1050, 1050, 1050),
  main_tx_start = 1000, main_tx_end = 40000,
  alt_tx_start = c(1050, 9000, 1050, 1050, 1050), alt_tx_end = 40000
)

res <- classify_consequence(switches)
res[, c("gene_name", "category", "subtype", "d_utr5_bp")]
```

## Calling the ORFs yourself

If you have transcript models and a genome, `orf_table()` splices each
transcript and calls its reading frame, producing the columns above.

```{r eval = FALSE}
orfs <- orf_table(models, genome = BSgenome.Hsapiens.UCSC.hg38::Hsapiens)
head(orfs[, c("tx_id", "aa_len", "utr5_len", "n_uATG")])
```

`find_orf()` is the piece that does the work, and is useful on its own:

```{r}
find_orf("GGGATGAAAGGGCCCTAAGGG", min_aa = 3)
```

The `n_uATG` count matters for the `longer` subtype in particular: an AUG gained
in a lengthened 5′ UTR can open an upstream reading frame and suppress
translation of the main one, which is what makes "longer 5′ UTR" a claim about
regulation rather than a curiosity.

## Drawing the composition

The shipped demo table holds 180 synthetic switches, classified by the real
decision tree.

```{r}
d <- utils::read.delim(
  system.file("extdata", "demo_consequence.tsv", package = "omakase")
)
consequence_summary(d, filter = both_full_length == 1)
```

The default figure is a radial donut. Each subtype gets an equal angular
sector, so a rare category is as legible as a common one, and the *radius*
carries the abundance:

```{r, fig.alt = "Radial donut of start-site switch consequences, each subtype an equal-angle sector whose radius encodes its abundance."}
plot_consequence(d, filter = both_full_length == 1,
                 title = "What the switch does to the transcript")
```

Three other styles are available when a ring is not what you want:

```{r fig.height = 3.2, fig.alt = "Lollipop chart of start-site switch consequences, one row per subtype, coloured by consequence category."}
plot_consequence(d, filter = both_full_length == 1, style = "lollipop")
```

```{r fig.height = 3.2, fig.alt = "Bar chart of start-site switch consequences, one bar per subtype, coloured by consequence category."}
plot_consequence(d, filter = both_full_length == 1, style = "bar")
```

Every style returns a `ggplot`, so the usual adjustments apply:

```{r fig.height = 3.2, fig.alt = "Lollipop chart of all switch consequences with a custom title and axis label."}
plot_consequence(d, style = "lollipop") +
  ggplot2::labs(title = "All switches", x = "Number of genes")
```
