---
title: "Getting started with omakase"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with omakase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.6, dpi = 120
)
```

```{r setup}
library(omakase)
```

## The shape of the package

omakase separates *counting* from *drawing*. Every reader returns the same
object, a `sashimi_data`, and every plotting function takes one. That is the
whole design:

```
BAM / CRAM  ─┐
rMATS events ┤
5'-tag BED   ├─→  sashimi_data  ─→  plot_sashimi()
STAR SJ.tab  ┤     (tidy tables)     sashimi_track()
tidy tables ─┘                       sashimi_annotation()
```

Because the intermediate is data rather than pixels, a slow pass over a large
BAM happens once, and re-plotting is instant. And because the plotting
functions return `ggplot` and `patchwork` objects, a figure can be adjusted
after it is drawn rather than by re-running everything with different flags.

## From alignments

The package ships six ENCODE libraries over the human *ABI1* locus, two from
each of three cell types.

```{r}
bams <- system.file("extdata", "samples.tsv", package = "omakase")
gtf  <- system.file("extdata", "annotation.gtf", package = "omakase")

read_manifest(bams)
```

The manifest is a plain TSV: identifier, path, then any number of metadata
columns. Column three groups the samples here, and `sashimi_from_bam()` reads
coverage and junctions over a region in one pass.

```{r}
sd <- sashimi_from_bam(
  bams,
  region     = "chr10:27040584-27048100",
  annotation = gtf,
  min_count  = 10
)
sd
```

The object is six tidy tables. Nothing is hidden:

```{r}
head(sd$junctions, 4)
```

Drawing it aggregates the two libraries of each cell type into one panel:

```{r}
plot_sashimi(sd, aggregate = "mean", arc_label_format = "count")
```

## Compressing introns

Most of a gene is intron, so a linear axis spends most of its width on empty
sequence. `shrink = TRUE` draws each intron at a reduced length while leaving
exons at scale 1, and the axis still reports true genomic coordinates.

```{r}
plot_sashimi(sd, aggregate = "mean", arc_label_format = "count",
             shrink = TRUE, arc_width_rule = "log", arc_width = 0.3)
```

The map itself is an object you can inspect and apply by hand:

```{r}
introns <- data.frame(start = c(27041000, 27044000), end = c(27043500, 27047500))
m <- intron_map(introns, 27040584, 27048100)
m
compress_coords(c(27040584, 27044000, 27048100), m)
```

It is exactly invertible, which is what lets the axis labels stay honest:

```{r}
expand_coords(compress_coords(27044000, m), m)
```

## Normalising

Neither ggsashimi nor rmats2sashimiplot normalises, so a deeply sequenced
library looks like a highly expressed gene. omakase makes the choice explicit:

```{r}
sd_cpm <- normalize_tracks(sd, "cpm",
                           library_sizes = c(ENCLB024ZZZ = 3.1e7,
                                             ENCLB025ZZZ = 2.8e7,
                                             ENCLB271TJH = 3.4e7,
                                             ENCLB459IUG = 2.9e7,
                                             ENCLB008ZZZ = 3.3e7,
                                             ENCLB009ZZZ = 3.0e7))
sd_cpm$meta$normalize_factors
```

`"size_factor"` uses DESeq2's median-of-ratios estimator and needs no library
sizes at all.

## Choosing the arc

Six geometries are available. The sine arc is the default; the x-spline is what
ggsashimi draws, the Bezier is MISO's.

```{r fig.height = 3}
library(patchwork)

wrap_plots(lapply(c("sine", "bezier", "elbow"), function(s) {
  sashimi_track(sd, group = "Endothelial", arc_shape = s,
                arc_label_format = "count") +
    ggplot2::ggtitle(s)
}), ncol = 1)
```

## Building the object yourself

If your counts come from somewhere omakase does not read, build the tables
directly. This is also how the package draws start-site activity figures, where
an "arc" is a pointer from a start site to the body of the transcript rather
than a counted junction.

```{r fig.height = 4}
pos <- seq(10000, 30000, length.out = 201)
wave <- function(scale) abs(sin(seq(0, 3 * pi, length.out = 201))) * scale

sd2 <- sashimi_data(
  loci = data.frame(locus_id = "g1", gene_name = "Demo", chrom = "chr12",
                    strand = "-", win_lo = 10000, win_hi = 30000,
                    main_apex = 12000, alt_apex = 18000),
  tracks = do.call(rbind, lapply(c("Early", "Late"), function(g) {
    data.frame(locus_id = "g1", group = g, pos = pos,
               value = wave(c(Early = 40, Late = 14)[[g]]))
  })),
  junctions = data.frame(
    locus_id = "g1", group = rep(c("Early", "Late"), each = 2),
    x0 = c(12000, 18000), x1 = 26000,
    count = c(103.01, 0.004, 12.4, 44.2),
    role = c("main", "alt")
  ),
  models = rbind(
    data.frame(locus_id = "g1", tx_id = "main", role = "main",
               start = c(11800, 20000, 25500), end = c(12400, 20600, 26200)),
    data.frame(locus_id = "g1", tx_id = "alt", role = "alt",
               start = c(17800, 20000, 25500), end = c(18300, 20600, 26200))
  )
)
sd2 <- compute_psi(sd2, main = "main", alt = "alt")
plot_sashimi(sd2)
```

Note the arc labels: `0.004` prints as `<0.01` rather than rounding to `0`.
`format_activity()` picks its precision from the magnitude of each value, so a
small non-zero number is never shown as absent.

## Where to go next

* `vignette("consequence")` — classifying and drawing what a start-site switch
  does to the transcript.
* `?plot_sashimi` — the full argument surface.
* `?sashimi_data` — the data contract, if you are writing your own reader.
